Complex Analysis of Askaryan Radiation: UHE-$\nu$ Identification and Reconstruction via the Hilbert Envelope of Observed Signals

### 2025-08-28 template.tex
fixed an equation, working on plots

\begin{abstract}
The detection of ultra-high energy neutrinos (UHE-$\nu$), with enegies above 10 PeV, has been a long-time goal in astroparticle physics.  Autonomous, radio-frequency (RF) UHE-$\nu$ detetectors have been deployed in polar regions.  These detectors rely on the Askaryan effect in ice for the neutrino signal.  The Askaryan effect occurs when the excess negative charge within a high-energy cascade radiates in a dense medium.  UHE-$\nu$ can induce such cascades that radiate in the RF bandwidth above thermal backgrounds.  To identify UHE-$\nu$ signals in future data from Askaryan-class detectors, analytic models of the Askaryan electromagnetic field have been created and compared to simulations and laboratory measurements.  These models have correctly described the Askaryan electromagnetic field, while leaving the effect of the RF detection channels on the field to simulation packages.  In this work, we present a fully analytic Askaryan model that accounts for the effect of an RF detection channel.  First, we derive formulas for the observed voltage trace, and for the Hilbert envelope of the trace.  Second, we match the analytic model to data computed for a detector with a single string of eight RF detection channels in NuRadioMC, a key Monte Carlo toolset in UHE-$\nu$ detection.  For UHE-$\nu$ signals above RF thermal backgrounds, we find correlation coefficients in excess of $0.95$.  Third, 92\% of UHE-$\nu$ signals pass a correlation threshold of 0.438, while all but 0.95 out of $1.57 \times 10^8$ RF thermal triggers are rejected.  At a thermal trigger rate of 1 Hz, this corresponds to 5 years of continuous operation.  Finally, we reconstruct the logarithm of the UHE-$\nu$ cascade energy.
\end{abstract}

@@

where $w(q)$ is the Faddeeva function.  Note that the Faddeeva function is \textit{entire}, meaning the $\Re\lbrace \rbrace$ operator and differentiation commute.  Inserting this result into Eq. \ref{eq:conv_sr}, and distributing the $\Re\lbrace \rbrace$ operator to the instances of $I(x,z)$, gives

@@

s * r = -\sqrt{\pi}R_0 E_0 \sigma_t^2 \\ \left(xe^{-x^2} \Re\left\lbrace w(q)\right\rbrace - \frac{1}{2}e^{-x^2} \frac{d\Re\lbrace w(q)\rbrace}{dx} \right)

@@

s * r = -\sqrt{\pi}R_0 E_0 \sigma_t^2 \\ \left(xe^{-x^2} \Re\left\lbrace w(q)\right\rbrace + \left(\frac{j}{2}\right)e^{-x^2} \frac{d\Re\lbrace w(q)\rbrace}{dq} \right) \label{eq:final3}

@@

\item A Monte Carlo data set was generated using NuRadioMC for UHE-$\nu$ interactions in a cylindrical ice volume with a depth of 0.65 km and a radius of 0.85 km.  The UHE-$\nu$ energies ranged from 10-100 PeV, with roughly equal numbers of UHE-$\nu$ in each 10 PeV bin.  The detector had one string of 8 RF dipole antennas.  Each RF channel had a filtered, amplified passpand of [80,1000] MHz, with 256 samples per channel at a 1 GHz sampling rate.  The RF trigger responded when 3 of the 8 voltage traces exceeded $\pm 3$ times the $v_{\rm rms}$ of the thermal noise within 256 ns.  For each signal, the Hilbert envelope of the coherently summed waveform (CSW) was calculated and correlated against Eqs. \ref{eq:final}-\ref{eq:final2}.  A quick optimization over the whole data set yielded 0.15 GHz and 0.025 GHz for the $f_0$ and $\gamma$ parameters, respectively.  Each correlation was maximized by varying the just the $\sigma_t$ parameter, yielding an optimized analytical envelope for each event.  A thermal noise waveform reflective of the channel properties was also generated for each event.  The noise waveform was also correlated against the optimized analytical envelope.  Both correlation distributions are shown in Fig. \ref{fig:fig3}.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,trim=3.25cm 8.25cm 4.5cm 9.0cm,clip=true]{Aug15_plot1.pdf}
\caption{\label{fig:fig3} (Black circles) Noise distribution. (Gray dashed line) Fitting function to noise distribution.  (Solid black line) Exponential fit to tail of noise distribution.  (Solid gray line) UHE-$\nu$ signal distribution.  (Dashed black line) Optimized correlation coefficient threshold value.}
\end{figure}

In Fig. \ref{fig:fig3}, the circles represent the normalized histogram of the correlation coefficent between the optimized analytic envelope and the thermal noise.  Functions of the forms $x^2 \exp(-0.5 x^2)$ and $\exp(-x)$ were fit to the noise distribution on the intervals [0.1,1.0] and [0.28,1], respectively.  The former is represented by the gray dashed line, and the latter is represented by the solid black line.  The solid gray line represents the signal distribution, which peaks at a correlation value of 0.96.  The vertical black dashed line represents a threshold of 0.438.  For the simulated UHE-$\nu$, 92\% of correlations between CSWs and optimized analytic envelopes are greater than or equal to this threshold.  Assuming the solid black line describes the noise distribution tail, with a noise trigger rate of 1 Hz, the threshold is equivalent to 0.95 noise events every 5 years.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,trim=3.25cm 8.25cm 4.5cm 9.0cm,clip=true]{Aug15_plot2.pdf}
\caption{\label{fig:fig4} The correlation versus SNR (dB) for UHE-$\nu$ signals (upper distribution) and RF thermal noise (lower distribution).  Color scale: normalized histogram value, with five equally spaced contours between 0.0 and 0.003.}
\end{figure}

@@

In Fig. \ref{fig:fig4}, the correlation coefficient is plotted versus the SNR in dB for the data shown in Fig. \ref{fig:fig3}.  The upper distribution corresponds to CSWs from UHE-$\nu$ events, and the flat distribution corresponds to the CSWs from RF thermal noise.  UHE-$\nu$ events with SNR values $>5$ dB tend to maximize the correlation coefficient, and the correlation coefficient decreases linearly with decreasing SNR (dB).  As expected, the correlation coefficient for thermal noise does not depend on SNR, since the thermal fluctuations that increase the SNR do not resemble the optimized analytic envelope.  Note that the SNR of a CSW does not equal the SNR of the individual voltage traces.  Rather, the individual voltage traces will have SNR values approximately 5-10 dB \textit{lower} than the CSW SNR.  If $N$ voltage traces contain signal, computing the CSW raises the linear SNR by a factor of $\sqrt{N}$, and adds to the SNR in dB a factor of $20\log_{10}(N)$.  For an event with 3 of 8 channels containing signal, $20\log_{10}(3)\approx 5$ dB, while $20\log_{10}(8)\approx 9$ dB.  The exact increase in SNR from traces to CSW depends on how many RF channels contain significant signal.

>>>>

\begin{abstract}
The detection of ultra-high energy neutrinos (UHE-$\nu$), with enegies above 10 PeV, has been a long-time goal in astroparticle physics.  Autonomous, radio-frequency (RF) UHE-$\nu$ detetectors have been deployed in polar regions.  These detectors rely on the Askaryan effect in ice for the neutrino signal.  The Askaryan effect occurs when the excess negative charge within a high-energy cascade radiates in a dense medium.  UHE-$\nu$ can induce such cascades that radiate in the RF bandwidth above thermal backgrounds.  To identify UHE-$\nu$ signals in future data from Askaryan-class detectors, analytic models of the Askaryan electromagnetic field have been created and compared to simulations and laboratory measurements.  These models have correctly described the Askaryan electromagnetic field, while leaving the effect of the RF detection channels on the field to simulation packages.  In this work, we present a fully analytic Askaryan model that accounts for the effect of an RF detection channel.  First, we derive formulas for the observed voltage trace, and for the Hilbert envelope of the trace.  Second, we match the analytic model to data computed for a detector with a single string of eight RF detection channels in NuRadioMC, a key Monte Carlo toolset in UHE-$\nu$ detection.  For UHE-$\nu$ signals above RF thermal backgrounds, we find correlation coefficients in excess of $0.95$.  Third, 96.9\% of UHE-$\nu$ signals pass a correlation threshold of 0.386, while all but 0.877 out of $1.57 \times 10^8$ RF thermal triggers are rejected.  At a thermal trigger rate of 1 Hz, this corresponds to 5 years of continuous operation.  Finally, we reconstruct the logarithm of the UHE-$\nu$ cascade energy.
\end{abstract}

@@

where $w(q)$ is the Faddeeva function.  Inserting this result into Eq. \ref{eq:conv_sr}, and distributing the $\Re\lbrace \rbrace$ operator to the instances of $I(x,z)$, gives

@@

s * r = -\sqrt{\pi}R_0 E_0 \sigma_t^2 \\ \Re\left\lbrace xe^{-x^2} w(q) - \frac{1}{2}e^{-x^2} \frac{d w(q)}{dx} \right\rbrace

@@

s * r = -\sqrt{\pi}R_0 E_0 \sigma_t^2 \\ \left(xe^{-x^2} \Re\left\lbrace w(q) \right\rbrace - \frac{1}{2}e^{-x^2} \Re\left\lbrace -j \frac{d w(q)}{dq} \right\rbrace \right) \label{eq:final3}

@@

\item A Monte Carlo data set was generated using NuRadioMC for UHE-$\nu$ interactions in a cylindrical ice volume with a depth of 0.65 km and a radius of 0.85 km.  The UHE-$\nu$ energies ranged from 10-100 PeV, with roughly equal numbers of UHE-$\nu$ in each 10 PeV bin.  The detector just a single string of 8 RF dipole antennas.  Each RF channel had a filtered, amplified passpand of [80,1000] MHz, with 256 samples per channel at a 1 GHz sampling rate.  The RF trigger responded when 3 of the 8 voltage traces exceeded $\pm 3$ times the $v_{\rm rms}$ of the thermal noise within 256 ns.  For each signal, the Hilbert envelope of the coherently summed waveform (CSW) was calculated and correlated against Eqs. \ref{eq:final}-\ref{eq:final2}.  A quick optimization over the whole data set yielded 0.15 GHz and 0.025 GHz for the $f_0$ and $\gamma$ parameters, respectively.  Each correlation was maximized by varying the just the $\sigma_t$ parameter, yielding an optimized analytical envelope for each event.  A thermal noise waveform reflective of the channel properties was also generated for each event.  The noise waveform was also correlated against the optimized analytical envelope.  Both correlation distributions are shown in Fig. \ref{fig:fig3}.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,trim=3.25cm 8.25cm 4.5cm 9.0cm,clip=true]{Aug15_plot1.pdf}
\caption{\label{fig:fig3} (Black circles) Noise distribution. (Gray dashed line) Fitting function to noise distribution.  (Solid gray line) UHE-$\nu$ signal distribution.  (Dashed black line) Optimized correlation coefficient threshold value.}
\end{figure}

In Fig. \ref{fig:fig3}, the circles represent the normalized histogram of the correlation coefficent between the optimized analytic envelope and thermal noise.  A fitting function of the form $x^2 \exp(-0.5 x^2)$ was fit to the noise distribution, and is represented by the gray dashed line.  The solid gray line represents the signal distribution, which peaks at a correlation value of 0.96.  Lower signal correlation values correspond to lower SNR values (Fig. \ref{fig:fig4}).  The vertical black dashed line represents a threshold of 0.386.  For the simulated UHE-$\nu$, 97\% of correlations between CSWs and optimized analytic envelopes are greater than or equal to this threshold.  Assuming the gray dashed line describes accurately the noise distribution tail, and assuming a thermal trigger rate of 1 Hz, the correlation threshold is equivalent to 0.88 noise events every 5 years.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,trim=3.25cm 8.25cm 4.5cm 9.0cm,clip=true]{Aug15_plot2.pdf}
\caption{\label{fig:fig4} The correlation versus SNR (dB) for UHE-$\nu$ signals (upper distribution) and RF thermal noise (lower distribution).  Color scale: normalized histogram value, with five equally spaced contours between 0.0 and 0.002.}
\end{figure}

@@

In Fig. \ref{fig:fig4}, the correlation coefficient is plotted versus the SNR in dB for the data shown in Fig. \ref{fig:fig3}.  The upper and lower distributions correspond to CSWs from UHE-$\nu$ events and thermal noise, respectively.  A linear relationship between the correlation coefficient for UHE-$\nu$ events and SNR$_{\rm dB}$ is observed, while the correlation coefficient for thermal noise is independent of SNR$_{\rm dB}$.  Note that the SNR of a CSW does not equal the SNR of the individual voltage traces.  Rather, the individual voltage traces will have SNR values approximately 5-10 dB \textit{lower} than the CSW SNR.  If $N$ voltage traces contain signal, computing the CSW raises the linear SNR by a factor of $\sqrt{N}$, and adds to the SNR in dB a factor of $20\log_{10}(N)$.  For an event with 3 of 8 channels containing signal, $20\log_{10}(3)\approx 5$ dB, while $20\log_{10}(8)\approx 9$ dB.  The exact increase in SNR from traces to CSW depends on how many RF channels contain significant signal.
